The Evolution of Phenotypically Invariant Genetic Networks

### 2017-06-08 handnotes.tex
edited discussion

\newcommand{\calA}{\mathcal{A}}
\newcommand{\diag}{\text{diag}}
\newcommand{\1}{\mathbbm{1}}

@@

  Define $V(\tau)$ as the parameterized change of coordinates matrix that preserves $B$ and $C$, with $\tau$ a vector of free parameters. The set of \emph{all} phenotypically invariant (minimal) gene networks is, 
  \begin{align}
    A(\tau) = V(\tau) A(0) V^{-1}(\tau) ,
  \end{align}
  and a \emph{Linear Evolutionary System} is, 
  \begin{align}
    \Sigma(\tau) = \left\{ \begin{array}{ll} \dot{\kappa}(t) &= A(\tau) \kappa(t) + B u(t) \\ \phi(t) &= C \kappa(t) \end{array} \right .
  \end{align}
  That is, all (linear and minimal) mechanisms capable of producing the same phenotype can be realized by a unique choice of $\tau$ in $\Sigma(\tau)$.

  More generally, \plr{introduce Kalman},
  we denote by $\calA_n(A_0)$ the set of all $n$-dimensional systems equivalent to $A_0$:
  \begin{align}
      \calA_n(A_0) = \left\{
          A \in R^{n \times n} : C e^{At} B = C e^{A_0 t} B \quad \text{for all}\; t \ge 0 
          \right\} .
  \end{align}
  Usually, the reference system $A_0$ is implicit and we write only $\calA$.

  \begin{example}[External Equivalence does not imply internal equivalence]
    Gene 1 dynamics (blue) are equivalent for network architectures $A(0)$ and $A(2)$, however the internal dynamics containing gene 2 (orange) are very different.
  \begin{figure}[H]
    \centering
    \includegraphics[width=0.5\textwidth]{osc_A0_A2_both_compare}
    %\caption{Gene 1 dynamics (blue) are equivalent for network architectures $A(0)$ and $A(2)$, however the internal dynamics containing gene 2 (orange) are very different.}
  \end{figure}
  \end{example}

  %%%%%%%%%%%%%%%%%%%%%%%%
  \section*{Genetic drift}

  In the absence of directional evolution by selective pressures,
  demographic and mutational noise
  will cause a population to move randomly on (or near) the space of phenotypically equivalent systems.
  On short to intermediate time scales this is thought to proceed mostly through genetic drift
  resampling available diversity.
  The amount and structure of 
  this standing variation is maintained on long time scales
  by mutation-selection balance
  shifts in the phenotypic optimum,
  diversifying selection, and/or other factors \citep{mut_sel_balance}.
  We will take a quantitative genetics framework to describe this process.

  If the system is a network of transcription factors,
  each row of the matrix $A$ represents the regulatory region of a particular transcription factor gene.
  Regulatory regions are typically kilobases in length,
  and so are expected to be represented in a population by a moderate number of haplotypes,
  within which both mutation and recombination generate new diversity at roughly equal rates.
  Nonetheless, as a first approximation we will assume that within the population,
  the distribution of additive genetic variation in each entry of $A$ is multivariate Gaussian
  with covariance matrix $G$.
  The state of the population is then determined by the population mean and covariance matrix.
  We then further decompose the population mean system into a position in the set of equivalent, optimal systems 
  (a projection onto the set $\calA$, which will usually be unique)
  and the deviation of the mean from that set.
  We would like to then consider these two components separately,
  since deviations away from the optima $\calA$ occur within populations,
  and so should be close to neutral,
  while two populations whose means drift in different directions along $\calA$
  may develop incompatibilities.

  Suppose that a large population with effective size $N_e$ has population mean $\bar A = A_0 + Z$,
  where $A_0$ is the closest point in $\calA$ and $Z$ is the deviation from $\calA$,
  and suppose the population varies about this point with covariance between $A_{ij}$ and $A_{k\ell}$
  equal to $G_{ij,k\ell}$.

  Treating $A$ as a vector, the mean per-generation displacement of $\bar A$ is approximately $G (G+E)^{-1} S$
  where $E$ is the covariance matrix of environmental noise affecting the phenotypes,

  It also seems reasonable to assume that population variation occurs on a smaller scale
  than the effects of selection -- 
  otherwise, a substantial portion of new offspring would die due to (recombination) genetic load each generation \citep{genetic_load}.

>>>>

\newcommand{\A}{\mathcal{A}}
\newcommand{\diag}{\text{diag}}
\newcommand{\1}{\mathbbm{1}}

@@

  Define $V(\tau)$ as the parameterized change of coordinates matrix that preserves $B$ and $C$, with $\tau$ a vector of free parameters. The set of \emph{all} phenotypically invariant (minimal) gene networks is, 
  \begin{align}
    A(\tau) = V(\tau) A(0) V^{-1}(\tau) ,
  \end{align}
  and a \emph{Linear Evolutionary System} is, 
  \begin{align}
    \Sigma(\tau) = \left\{ \begin{array}{ll} \dot{\kappa}(t) &= A(\tau) \kappa(t) + B u(t) \\ \phi(t) &= C \kappa(t) \end{array} \right .
  \end{align}
  That is, all (linear and minimal) mechanisms capable of producing the same phenotype can be realized by a unique choice of $\tau$ in $\Sigma(\tau)$.
  \begin{example}[External Equivalence does not imply internal equivalence]
    Gene 1 dynamics (blue) are equivalent for network architectures $A(0)$ and $A(2)$, however the internal dynamics containing gene 2 (orange) are very different.
  \begin{figure}[H]
    \centering
    \includegraphics[width=0.5\textwidth]{osc_A0_A2_both_compare}
    %\caption{Gene 1 dynamics (blue) are equivalent for network architectures $A(0)$ and $A(2)$, however the internal dynamics containing gene 2 (orange) are very different.}
  \end{figure}
  \end{example}
  Evolution thus proceeds as a random walk in phenotypically invariant network space.
